---
title: "Factorial ANOVA Demo"
author: "Christopher J. Schmank, PhD"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Factorial ANOVA | Research Context Prompt

A local dating expert was interested in studying first-date satisfaction and whether it depended on variables like (1) date attire (i.e., Athletic Wear, Leisure Wear, or Dressed Up) and (2) dating application used (OKCupid, Tinder).

Participants consented to be in their study after setting up their first date using either OKCupid or Tinder (this variable could not be randomly assigned). Upon meeting inclusion criteria and joining the study, the dating expert randomly assigned each participant to one of the three attire conditions.

The research question was: Does first-date satisfaction differ interactively when considering dating attire and the mobile application used?

#### 3 x 2 Factorial ANOVA | Data

```{r message = FALSE, warning = FALSE}
library(psych)
library(tidyverse)
library(jmv)
library(ggpubr)
library(ggrain)
library(apaTables)
library(ez)
library(rstatix)

dat_date <- read.csv("FACTORIAL_DEMO.csv")

# Set IVs to factors using as.factor() or factor() 
dat_date$Application<-as.factor(dat_date$Application)
dat_date$Outfit<-as.factor(dat_date$Outfit)

# Assess number of participants and variables in data set
dim(dat_date)
head(dat_date,4)
```

#### Factorial ANOVA | Descriptive Stats

```{r}

# Descriptive Stats
describeBy(dat_date$Satisfaction, 
           list(dat_date$Outfit,dat_date$Application),       
           mat=TRUE)              

```


#### Factorial ANOVA | Histogram

```{r}

ggplot(data = dat_date,
       mapping = aes(x = Satisfaction)) +
  geom_histogram(bins = 8, color = "steelblue") +
  labs(y = "Count", x = "Satisfaction")+
  facet_wrap(dat_date$Application,ncol=1)

```

```{r}

ggplot(data = dat_date,
       mapping = aes(x = Satisfaction)) +
  geom_histogram(bins = 8, color = "steelblue") +
  labs(y = "Count", x = "Satisfaction")+
  facet_wrap(dat_date$Outfit,ncol=1)

```

```{r}

ggplot(data = dat_date,
       mapping = aes(x = Satisfaction)) +
  geom_histogram(bins = 9, color = "steelblue") +
  labs(y = "Count",
       x = "Satisfaction")+
  facet_wrap(~dat_date$Outfit+dat_date$Application,ncol = 2)

```

#### Factorial ANOVA | Boxplots and Raincloud Visualizations

```{r}

ggplot(data = dat_date,
       mapping = aes(y = Satisfaction, x = Outfit)) +
  geom_boxplot() +
    labs(y = "Date Satisfaction")
```

```{r}

ggplot(data = dat_date,
       mapping = aes(y = Satisfaction, x = Application)) +
  geom_boxplot() +
    labs(y = "Date Satisfaction")

```

```{r}

ggplot(data = dat_date,
       mapping = aes(y = Satisfaction, x = Outfit)) +
  geom_boxplot() +
  facet_wrap(~Application)+
    labs(y = "Date Satisfaction")

```

```{r}

ggplot(data = dat_date,
       mapping = aes(y = Satisfaction, x = Application)) +
  geom_boxplot() +
  facet_wrap(~Outfit) +
    labs(y = "Date Satisfaction")

```

```{r}
ggplot(data = dat_date,
       aes(x = Outfit, y = Satisfaction, 
           fill = Application))+
  geom_rain(show.legend=FALSE) +
  facet_wrap(~Application)+
  scale_colour_grey(aesthetics = "fill",
                    start = .6,
                    end = 1) +
  coord_flip() +
  ylab("Date Satisfaction")+
  xlab("Outfit")+
  theme_classic()
```

#### Factorial ANOVA | ANOVA() Omnibus Assumption Checks

```{r}

# Omnibus Assumption Checking
ANOVA(data = dat_date, 
      dep = 'Satisfaction',             
      factors = c('Application','Outfit'),
      homo = TRUE,                      
      norm = TRUE,                      
      qq = TRUE)      

```

#### Factorial ANOVA | Group Level Assumption Checks

```{r}

dat_date %>%                    
  group_by(Application,Outfit) %>%  
  shapiro_test(Satisfaction)    

```

#### Factorial ANOVA | Group Level Assumption Checks

```{r}

ggqqplot(dat_date,                   
         'Satisfaction',             
         facet.by = c("Application","Outfit"))  

```

#### Factorial ANOVA | Conducting the ANOVA

```{r}
options(digits = 3)
ANOVA(data = dat_date,                    
      dep = 'Satisfaction',               
      factors = list('Application','Outfit'),  
      effectSize = 'partEta',             
      postHoc = ~Outfit + Outfit:Application, 
      postHocCorr = 'bonf',
      postHocES = 'd',   
      postHocEsCi = TRUE,
      emMeans = ~Application + Outfit + Outfit:Application,         
      emmPlots = FALSE,
      emmPlotData = TRUE,
      emmTables = TRUE)                   

```

#### Factorial ANOVA | ANOVA Output: Interaction Simple Effects Code

```{r}
model <- lm(Satisfaction~Application*Outfit, dat_date)
# Effect of Outfit across each Application
# When statistically significant interaction occurs we assess simple effects
one.way <- dat_date %>%                 
  group_by(Outfit) %>%                  
  anova_test(dv = Satisfaction,         
             between = Application,
             error = model)%>%  
  get_anova_table() 
one.way 

one.way2 <- dat_date %>%
  group_by(Application) %>%
  anova_test(dv = Satisfaction,
             between = Outfit,
             error = model)%>%
  get_anova_table()
one.way2

```

The simple main effect of Outfit on date satisfaction was statistically significant for both OKCupid and Tinder users (*p* < .001). In other words, there was a statistically significant difference in mean date satisfaction scores between OKCupid users based on date attire [*F*(2,114) = 37.07, *p* < .001]. And the same conclusion holds true for Tinder users [*F*(2,114) = 33.07, *p* < .001]. We would need to do further post-hoc testing to figure out where the differences occur.  

**BUT** this is not as informative as looking at the simple main effects of Application on date satisfaction for Athletic, Dressed Up, and Leisure Wear conditions. There was no statistically significant difference between applications when participants wore leisure wear [*F*(1,114) = 0.19, *p* > .05]. But significant differences arose across applications for those wearing Athletic Wear [*F*(1,114) = 30.11, *p* < .001, $\eta_p^2$ = .21) and those that Dressed Up [*F*(1,114) = 35.05, *p* < .001, $\eta_p^2$ = .24]. 

#### Factorial ANOVA | Professional ANOVA Visualization Code

```{r}

pwc1 <- dat_date %>%
  group_by(Application) %>%                               
  emmeans_test(Satisfaction~Outfit,                 
               p.adjust.method = "bonferroni") %>%   
  add_xy_position(x = "Application") 

ggboxplot(dat_date,                            
          x = "Application",                       
          y = "Satisfaction",                 
          color = "Outfit") +               
  stat_pvalue_manual(pwc1,                     
                     hide.ns = FALSE,
                     tip.length = 0) +       
  labs(y = "Date Satisfaction",
       x = "Application Used")
```
```{r}
pwc2 <- dat_date %>%
  group_by(Outfit) %>%
  emmeans_test(Satisfaction~Application,
               p.adjust.method = "bonferroni") %>%
  add_xy_position(x = "Outfit")

ggboxplot(dat_date,                            
          x = "Outfit",                       
          y = "Satisfaction",                 
          color = "Application") +               
  stat_pvalue_manual(pwc2,                     
                     hide.ns = FALSE,
                     tip.length = 0) +       
  labs(                                       
    caption = get_pwc_label(pwc2),            
    y = "Date Satisfaction",
    x = "Outfit Worn")
```

#### Factorial ANOVA | APA Style ANOVA Tables

```{r}
# ANOVA Source Table
aovm3<-ANOVA(data = dat_date,
             dep = 'Satisfaction',
             factors = list('Application','Outfit'),
             effectSize = 'partEta')$model

apa.aov.table(aovm3,
              table.number=5,
              conf.level=.95)

#Generate Marginal Means Table
apa.2way.table(Outfit,
               Application,                      
               Satisfaction,
               dat_date,    
               table.number=6,
               show.conf.interval = TRUE)
```