---
title: "Mixed-Factorial ANOVA Demo"
author: "Christopher J. Schmank, PhD"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Mixed Factorial ANOVA | Research Context Prompt

The same stress researcher from our previous demonstration conducted another study to determine the impact of psychosocial stress on participants processing speed ability using a mixed-factorial research design. 

The researcher randomly assigned 30 participants to one of two stress conditions: (1) Low Stress and (2) High Stress. Participants were assessed at three separate times across their academic career (Sophomore, Junior, and Senior) measuring a new processing speed performance measure. The dependent measure in this task was processing speed ability measured in milliseconds (500 ms = .5 seconds).

Again, this researcher believed that regulating feelings of stress would impair concurrent performance of processing speed and that these effects would materialize across time.

#### 3 x 2 Mixed Factorial ANOVA | Data

```{r message = FALSE, warning = FALSE}

library(psych)
library(tidyverse)
library(jmv)
library(ggpubr)
library(ggrain)
library(apaTables)
library(ez)
library(rstatix)

dat_MF <- read.csv("Mixed_Factorial.csv")

# SUPER IMPORTANT FOR WITHIN SUBJECT DESIGNS
# Set Subject or ID variable to a factor  
dat_MF$Subject <- factor(dat_MF$Subject)

# Set IVs to factors
dat_MF$Condition <- factor(dat_MF$Condition, labels = c("Low-Stress","High-Stress"))
dat_MF$Rank <- factor(dat_MF$Rank, labels = c("Senior", "Junior", "Sophomore"))

# Assess number of participants and variables in data set
dim(dat_MF)
head(dat_MF)
```

```{r}
# To use anovaRM() we must reshape our data set from `long` to `wide` 
dat_MFW <- dat_MF %>% 
  pivot_wider(names_from = Rank,
              values_from = Time)

# Let's Calculate an Average Time variable for assumption checking
dat_MFW$averageTime <- (dat_MFW$Sophomore + dat_MFW$Junior + dat_MFW$Senior) / 3
```

#### Mixed Factorial ANOVA | Descriptive Stats

```{r}

# Descriptive Stats
describeBy(dat_MF$Time,
           list(dat_MF$Condition, dat_MF$Rank),
           mat=TRUE)

```


#### Mixed Factorial ANOVA | Histogram

```{r}

ggplot(data = dat_MF,
       mapping = aes(x = Time)) +
  geom_histogram(bins = 8, color = "steelblue") +
  labs(y = "Count", x = "Time")+
  facet_wrap(~dat_MF$Rank, ncol =1)

```

```{r}

ggplot(data = dat_MF,
       mapping = aes(x = Time)) +
  geom_histogram(bins = 8, color = "steelblue") +
  labs(y = "Count", x = "Time")+
  facet_wrap(~dat_MF$Condition, ncol=1)

```

```{r}

ggplot(data = dat_MF,
       mapping = aes(x = Time)) +
  geom_histogram(bins = 8, color = "steelblue") +
  labs(y = "Count", x = "Time")+
  facet_wrap(~dat_MF$Rank + dat_MF$Condition, ncol=2)

```
#### Mixed Factorial ANOVA | Boxplot and Raincloud Plot Visualiztaions

```{r }

ggplot(data = dat_MF,
       mapping = aes(y = Time, x = Condition)) +
  geom_boxplot() +
    labs(y = "Processing Speed (Milliseconds)")

```

```{r}

ggplot(data = dat_MF,
       mapping = aes(y = Time, x = Rank)) +
  geom_boxplot() +
    labs(y = "Processing Speed (Milliseconds)")

```

```{r}

ggplot(data = dat_MF,
       mapping = aes(y = Time, x = Condition)) +
  geom_boxplot() +
  facet_wrap(~Rank)+
    labs(y = "Processing Speed (Milliseconds)")

```

```{r}

ggplot(data = dat_MF,
       mapping = aes(y = Time, x = Rank)) +
  geom_boxplot() +
  facet_wrap(~Condition) +
    labs(y = "Processing Speed (Milliseconds)")

```

```{r}
ggplot(data = dat_MF,
       aes(x = Rank, y = Time, 
           fill = Condition))+
  geom_rain(show.legend=FALSE) +
  facet_wrap(~Condition)+
  scale_colour_grey(aesthetics = "fill",
                    start = .6,
                    end = 1) +
  coord_flip() +
  ylab("Processing Speed (Milliseconds)")+
  theme_classic()
```

#### Mixed Factorial ANOVA | anovaRM() Omnibus Assumption Checks

```{r}

anovaRM(data=dat_MFW,                          
        bs = list('Condition'),             
        bsTerms = list('Condition'),
        rm = list(                  
             list(label = 'Rank',    
                  levels = list('Sophomore',    
                                'Junior',
                                'Senior'))),  
        rmCells = list(
           list(measure = 'Sophomore',
                cell = 'Sophomore'),
           list(measure = 'Junior',
                cell = 'Junior'),
           list(measure = 'Senior',
                cell = 'Senior')),             
        rmTerms= list('Rank'),                
        leveneTest = TRUE,                   
        qq = TRUE,                           
        spherTests = TRUE)                   

```

#### Mixed Factorial ANOVA | Group Level Assumption Checks

```{r}

dat_MF %>% 
  group_by(Condition, Rank) %>% 
  shapiro_test(Time)

```

#### Mixed Factorial ANOVA | Group Level Assumption Checks

```{r}

ggqqplot(dat_MF,
         'Time',
         facet.by = c("Condition","Rank"))

```

#### Mixed Factorial ANOVA | Omnibus Levene's Test Assumption Checks

```{r}

dat_MFW %>% 
  levene_test(averageTime ~ Condition,   
              center = mean)

```

#### Mixed Factorial ANOVA | Group Level Assumption Checks

```{r}

# Must be conducted ON A MATRIX!! Also, note that it is using the long version of the data!

box_m(as.data.frame(dat_MF$Time),      
      dat_MF$Condition)        

```

#### Mixed Factorial ANOVA | Conducting the ANOVA

```{r}
options(digits = 3)
anovaRM(data = dat_MFW,
        bs = list('Condition'),             
        bsTerms = list('Condition'),
        rm = list(                  
             list(label = 'Rank',    
                  levels = list('Sophomore',    
                                'Junior',
                                'Senior'))),  
        rmCells = list(
           list(measure = 'Sophomore',
                cell = 'Sophomore'),
           list(measure = 'Junior',
                cell = 'Junior'),
           list(measure = 'Senior',
                cell = 'Senior')),             
        rmTerms = list('Rank'),
        spherCorr = c("none","HF"),
        postHoc = list('Rank', c('Condition','Rank')),
        postHocCorr = 'bonf',  
        emMeans = list('Rank','Condition', c('Rank','Condition')), 
        emmPlots = FALSE,
        emmTables = TRUE,
        effectSize = 'partEta')
```

#### Mixed Factorial ANOVA | ANOVA Output: Interaction Simple Effects

```{r}

one.way <- dat_MF %>%                  
  group_by(Condition) %>%              
  anova_test(dv = Time,                
             wid = Subject,            
             within = Rank,
             effect.size = 'ges') %>%  
  get_anova_table(correction = "none") %>%
  adjust_pvalue(method = "bonferroni")

one.way


one.way2 <- dat_MF %>%
  group_by(Rank) %>%
  anova_test(dv = Time,
             between = Condition,
             effect.size = 'ges') %>%
  get_anova_table() %>%
  adjust_pvalue(method = "bonferroni")

one.way2


```

Demonstrates that there are significant differences between the Ranks for participants in the Low-Stress [*F*(2,28) = 86.50, *p* < .001, $\eta^2$ = .47] and High-Stress [*F*(2,28) = 83.79, *p* < .001, $\eta^2$ = .77] conditions---furthermore, there is an issue of Sphericity demonstrated for the High-Stress condition (note the degrees of freedom!). However, this requires that we conduct additional post hoc tests.   

On the other hand, based on Bonferroni corrected p-values there were no statistically significant differences between stress conditions for students regardless of school rank: sophomore year [*F*(1,28) = 5.00, *p* = .10, $\eta_p^2$ = .15], junior year [*F*(1,28) = 0.31, *p* = .58, $\eta_p^2$ = .01], and senior year [*F*(1,28) = 4.89, *p* = .11, $\eta_p^2$ = .15]. The lack of statistical significance here is probably due to the sample size being under-powered, look at the partial eta-square values.


#### Mixed Factorial ANOVA | Professional ANOVA Visualization Code

```{r}
pwc1 <- dat_MF %>%
  group_by(Rank) %>%
  pairwise_t_test(Time~Condition,
                  p.adjust.method = "bonferroni") %>%
  add_xy_position(x = "Rank", 
                  group = "Condition")  

ggboxplot(dat_MF,                                  
          x = "Rank",                              
          y = "Time",                              
          color = "Condition",
          palette = "grey")+                     
  stat_pvalue_manual(pwc1,                          
                     hide.ns = FALSE,
                     tip.length = 0)
```

```{r}
pwc2 <- dat_MF %>%
  group_by(Condition) %>%
  pairwise_t_test(Time~Rank,
                  p.adjust.method = "bonferroni",
                  paired = TRUE) %>%
  add_xy_position(x = "Condition", 
                  group = "Rank")  

ggboxplot(dat_MF,                                  
          x = "Condition",                              
          y = "Time",                              
          color = "Rank",
          palette = "grey")+                     
  stat_pvalue_manual(pwc2,                          
                     hide.ns = FALSE,
                     tip.length = 0)

```

#### Mixed Factorial ANOVA | APA Style ANOVA Tables

```{r}

res.aov <- dat_MF %>%
  ezANOVA(dv = .(Time),
          wid = .(Subject), 
          between= .(Condition), 
          within = .(Rank), 
          detailed = TRUE)

res.aov$ANOVA$ges <- res.aov$ANOVA$SSn/(res.aov$ANOVA$SSn+res.aov$ANOVA$SSd)

res.aov$ANOVA<-res.aov$ANOVA[c(1:3,6:7,9)]


apa.ezANOVA.table(res.aov,                               
                  table.title = "Mixed Factorial ANOVA",
                  correction="HF",
                  table.number=7)



#Generate Marginal Means Table
apa.2way.table(Rank,                          
               Condition,                               
               Time,                               
               dat_MF,                             
               table.number=8,
               show.conf.interval = TRUE)
```