---
title: "Mixed-Factorial ANOVA Demo"
author: "Professor Christopher J. Schmank"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Mixed Factorial ANOVA | Research Context Prompt

The same stress researcher from our previous demonstration conducted another study to determine the impact of psychosocial stress on participants processing speed ability using a mixed-factorial research design. 

The researcher randomly assigned 30 participants to one of two stress conditions: (1) Low Stress and (2) High Stress. Participants were assessed at three separate times across their academic career (Sophomore, Junior, and Senior) measuring a new processing speed performance measure. The dependent measure in this task was processing speed ability measured in milliseconds (500 ms = .5 seconds).

Again, this researcher believed that regulating feelings of stress would impair concurrent performance of processing speed and that these effects would materialize across time.

#### Mixed Factorial ANOVA | Data

```{r message = FALSE, warning = FALSE}

library(psych)
library(tidyverse)
library(jmv)
library(ggpubr)
library(apaTables)
library(ez)
library(rstatix)
library(reshape)

dat_MF <- read.csv("Mixed_Factorial.csv")
dat_MF$Time <- 1000 - dat_MF$Time

# SUPER IMPORTANT FOR WITHIN SUBJECT DESIGNS
# Set Subject or ID variable to a factor  
dat_MF$Subject <- factor(dat_MF$Subject)

# Set IVs to factors  
dat_MF$Condition <- factor(dat_MF$Condition, labels = c("Low-Stress","High-Stress"))
dat_MF$Rank <- factor(dat_MF$Rank, labels = c("Sophomore", "Junior", "Senior"))

# Assess number of participants and variables in data set
dim(dat_MF)
head(dat_MF)

# To use anovaRM() we must reshape our data set from `long` to `wide` 
dat_MFW <- cast(dat_MF,
                Subject + Condition ~ Rank, #Isolates each ID and Condition to form columns for Rank
                value = "Time")             #Tells function to use Time variable to fill in Ranks

dat_MFW$averageTime <- (dat_MFW$Sophomore + dat_MFW$Junior + dat_MFW$Senior) / 3 # This is used later for assumption checking
```

#### Mixed Factorial ANOVA | Descriptive Stats

```{r}

# Descriptive Stats
describeBy(dat_MF$Time,
           list(dat_MF$Condition, dat_MF$Rank),
           mat=TRUE)

```


#### Mixed Factorial ANOVA | Histogram

```{r}

ggplot(data = dat_MF,
       mapping = aes(x = Time)) +
  geom_histogram(bins = 10, color = "steelblue") +
  labs(y = "Count", x = "Time")+
  facet_wrap(~dat_MF$Condition + dat_MF$Rank)

```
#### Mixed Factorial ANOVA | Boxplot

```{r }

ggplot(data = dat_MF,
       mapping = aes(y = Time, x = Condition)) +
  geom_boxplot() +
    labs(y = "Processing Speed (Milliseconds)")

```

#### Mixed Factorial ANOVA | Boxplot

```{r}

ggplot(data = dat_MF,
       mapping = aes(y = Time, x = Rank)) +
  geom_boxplot() +
    labs(y = "Processing Speed (Milliseconds)")

```

#### Mixed Factorial ANOVA | Violin Plot

```{r}

ggplot(data = dat_MF,
       aes(x = Condition, y = Time, 
           fill = Condition))+
  geom_violin(show.legend=FALSE)+
  ylab("Processing Speed (Milliseconds)")+
  xlab("Condition")+
  scale_fill_manual(values=c("#D55e00","#56B4E9","#009E73"))

```

#### Mixed Factorial ANOVA | Violin Plot

```{r}

ggplot(data = dat_MF,
       aes(x = Rank, y = Time, 
           fill = Rank))+
  geom_violin(show.legend=FALSE)+
  ylab("Processing Speed (Milliseconds)")+
  xlab("Rank")+
  scale_fill_manual(values=c("#D55e00","#56B4E9","#009E73"))

```

#### Mixed Factorial ANOVA | Grouped Violin Plot

```{r}

ggplot(data = dat_MF,
       aes(x = Rank, y = Time, 
           fill = Condition))+
  geom_violin(show.legend=TRUE)+
  ylab("Processing Speed (Milliseconds)")+
  xlab("Rank")+
  scale_fill_manual(values=c("#D55e00","#56B4E9","#009E73"))

```

#### Mixed Factorial ANOVA | anovaRM() Omnibus Assumption Checks

```{r}

anovaRM(data=dat_MFW,                          
        bs = list('Condition'),             
        bsTerms = list('Condition'),
        rm = list(                  
             list(label = 'Rank',    
                  levels = list('Sophomore',    
                                'Junior',
                                'Senior'))),  
        rmCells = list(
           list(measure = 'Sophomore',
                cell = 'Sophomore'),
           list(measure = 'Junior',
                cell = 'Junior'),
           list(measure = 'Senior',
                cell = 'Senior')),             
        rmTerms= list('Rank'),                
        leveneTest = TRUE,                   
        qq = TRUE,                           
        spherTests = TRUE)                   

```

#### Mixed Factorial ANOVA | Group Level Assumption Checks

```{r}

dat_MF %>% 
  group_by(Condition, Rank) %>% 
  shapiro_test(Time)

```

#### Mixed Factorial ANOVA | Group Level Assumption Checks

```{r}

ggqqplot(dat_MF,
         'Time',
         facet.by = c("Condition","Rank"))

```

#### Mixed Factorial ANOVA | Group Level Assumption Checks

```{r}

dat_MFW %>% 
  levene_test(averageTime ~ Condition,   
              center = mean)

```

#### Mixed Factorial ANOVA | Group Level Assumption Checks

```{r}

# Must be conducted ON A MATRIX!! Also, note that it is using the long version of the data!

box_m(as.data.frame(dat_MF$Time),      
      dat_MF$Condition)        

```

#### Mixed Factorial ANOVA | Conducting the ANOVA

```{r}
options(digits = 3)
anovaRM(data = dat_MFW,
        bs = list('Condition'),             
        bsTerms = list('Condition'),
        rm = list(                  
             list(label = 'Rank',    
                  levels = list('Sophomore',    
                                'Junior',
                                'Senior'))),  
        rmCells = list(
           list(measure = 'Sophomore',
                cell = 'Sophomore'),
           list(measure = 'Junior',
                cell = 'Junior'),
           list(measure = 'Senior',
                cell = 'Senior')),             
        rmTerms = list('Rank'),
        postHoc = list('Rank', c('Rank','Condition')),
        postHocCorr = 'bonf',  
        emMeans = list('Rank','Condition', c('Rank','Condition')), 
        emmPlots = TRUE,
        emmTables = TRUE,
        effectSize = 'partEta')
```

#### Mixed Factorial ANOVA | ANOVA Output: Interaction Simple Effects

```{r}

one.way <- dat_MF %>%                  
  group_by(Condition) %>%              
  anova_test(dv = Time,                
             wid = Subject,            
             within = Rank,            
             effect.size = 'ges') %>%  
  get_anova_table() %>%                
  adjust_pvalue(method = "bonferroni") 

one.way


one.way2 <- dat_MF %>%                  
  group_by(Rank) %>%              
  anova_test(dv = Time,                
             between = Condition,            
             effect.size = 'ges') %>%  
  get_anova_table() %>%                
  adjust_pvalue(method = "bonferroni") 

one.way2


```

Demonstrates that there are significant differences between the Ranks for participants in the Low-Stress [*F*(2, 28) = 86.50, *p* < .001] and High-Stress [*F*(1.33, 18.64) = 83.79, *p* < .001] conditions---furthermore, there is an issue of Sphericity demonstrated for the High-Stress condition (note the degrees of freedom!).


#### Mixed Factorial ANOVA | Professional ANOVA Visualization Code

```{r }

# Generates Marginal Means for Visualization--the p-values here are more appropriate to interpret compared to the chunk above!
pair <- dat_MF %>%
  group_by(Rank) %>%
  pairwise_t_test(Time~Condition,
                  p.adjust.method = "bonferroni") %>%
  add_xy_position(x = "Rank", 
                  group = "Condition")  

pair

pair2 <- dat_MF %>%
  group_by(Condition) %>%
  pairwise_t_test(Time~Rank,
                  p.adjust.method = "bonferroni") %>%
  add_xy_position(x = "Condition", 
                  group = "Rank")  

pair2

```

Demonstrates that the difference between high- and low-stress conditions only materialized when participants were in their Sophomore (*p* = .04) and Senior (*p* = .03) years, but not when they were Juniors (*p* = .58)

#### Mixed Factorial ANOVA | Professional ANOVA Visualization Code

```{r}

ggboxplot(dat_MF,                                  
          x = "Rank",                              
          y = "Time",                              
          color = "Condition",
          palette = "grey")+                     
  stat_pvalue_manual(pair,                          
                     hide.ns = FALSE,
                     tip.length = 0)+             
  labs(                                             
    caption = get_pwc_label(pair)                   
  )

ggboxplot(dat_MF,                                  
          x = "Condition",                              
          y = "Time",                              
          color = "Rank",
          palette = "grey")+                     
  stat_pvalue_manual(pair2,                          
                     hide.ns = FALSE,
                     tip.length = 0)+             
  labs(                                             
    caption = get_pwc_label(pair2)                   
  )

```

#### Mixed Factorial ANOVA | APA Style ANOVA Tables

```{r}

res.aov <- dat_MF %>% ezANOVA(dv = .(Time), 
                                wid = .(Subject), 
                                between= Condition, 
                                within = .(Rank), 
                                detailed = TRUE)

res.aov$ANOVA$ges <- res.aov$ANOVA$SSn/(res.aov$ANOVA$SSn+res.aov$ANOVA$SSd)

res.aov$ANOVA<-res.aov$ANOVA[c(1:3,6:7,9)]


apa.ezANOVA.table(res.aov,                               
                  table.title = "Mixed Factorial ANOVA",
                  correction="GG",
                  filename = "This is my file.doc",
                  table.number=7)



#Generate Marginal Means Table
apa.2way.table(Rank,                          
               Condition,                               
               Time,                               
               dat_MF,                             
               table.number=8,
               show.conf.interval = TRUE)
```