---
title: "One-Way Within-Subjects ANOVA Demo"
author: "Professor Christopher J. Schmank"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### One-way Within-Subjects ANOVA | Research Context Prompt

A local stress researcher is interested in the relationship between psychosocial stress levels and processing speed. After conducting a brief literature review they have hypothesized that regulating feelings of stress impairs concurrent performance of processing speed.

The researcher conducted a study where psychosocial stress was manipulated into three conditions: low stress, mild stress, and strong stress. Money was tight and recruitment was difficult, so they implemented a within-subjects, repeated measures design. Data was collected from 20 participants that consented to participate. The dependent measure in this task was processing speed ability measured in 10-second intervals (i.e., 1.3 = 13 seconds!!)

#### One-way Within-Subjects ANOVA | Data

```{r message = FALSE, warning = FALSE}

library(psych)
library(tidyverse)
library(jmv)
library(ggpubr)
library(apaTables)
library(ez)
library(rstatix)

dat_stress <- read.csv("RM.ANOVA.Demo.csv")
dim(dat_stress)
head(dat_stress)

```

#### One-way Within-Subjects ANOVA | Descriptive Stats

```{r}

# Descriptive Stats
describe(dat_stress[,2:4])

```

#### One-way Within-Subjects ANOVA | Boxplot

```{r}

dat_stressL<- dat_stress %>% 
  gather(key="Stress",
         value="Score",
         Calm,
         Mild,
         Strong)

dat_stressL$Stress<-factor(dat_stressL$Stress,labels = c("Calm","Mild","Strong"))

ggplot(data = dat_stressL,
       mapping = aes(y = Score, x = Stress)) +
  geom_boxplot() +
    labs(y = "Processing Speed (10-second Intervals)")

```

#### One-way Within-Subjects ANOVA | Omnibus Normality Check

```{r}

anovaRM(data=dat_stress,
        rm = list(
          list(label = 'Stress',levels = list('Calm','Mild','Strong'))),
        rmCells = list(
          list(measure = 'Calm', cell = 'Calm'),
          list(measure = 'Mild', cell = 'Mild'),
          list(measure = 'Strong', cell = 'Strong')),
        rmTerms= list('Stress'),
        qq = TRUE,
        spherTests = TRUE)

```

#### One-way Within-Subjects ANOVA | Normality Assumption Check

```{r}

ggqqplot(dat_stressL,
         'Score',
         facet.by = "Stress")

```


#### One-way Within-Subjects ANOVA | Conducting the ANOVA 

```{r}
options(digits = 3)
anovaRM(data=dat_stress,
        rm = list(
          list(label = 'Stress',levels = list('Calm','Mild','Strong'))),
        rmCells = list(
          list(measure = 'Calm', cell = 'Calm'),
          list(measure = 'Mild', cell = 'Mild'),
          list(measure = 'Strong', cell = 'Strong')),
        rmTerms= list('Stress'),
        effectSize = 'eta',
        postHoc='Stress',
        postHocCorr = list("bonf"),
        emMeans=list(
          list('Stress')),
        emmTables=TRUE,
        qq = TRUE,
        spherTests = TRUE,
        spherCorr = list('none', 'GG'))

```

#### One-way Within-Subjects ANOVA | Conducting the ANOVA

```{r}

res <- anova_test(dat_stressL,
                  dv=Score,
                  wid=ID,
                  within=Stress,
                  effect.size = "ges",
                  type = 3,
                  detailed=TRUE)

get_anova_table(res)

```

#### One-way Within-Subjects ANOVA | ANOVA Output: Manual Post-Hoc Tests Code

```{r}

pair<-dat_stressL %>%
  pairwise_t_test(Score~Stress,
                  paired=TRUE, 
                  p.adjust.method = "bonferroni") %>%
  add_xy_position(x = "Stress")

```

#### One-way Within-Subjects ANOVA | Professional ANOVA Visualization Code

```{r}

ggboxplot(dat_stressL, 
          x = "Stress", 
          y = "Score") +
  stat_pvalue_manual(pair,
                     hide.ns = TRUE, 
                     tip.length = .01) +
  labs(y = "Processing Speed (10-second Intervals)", x = "Stress Level")

```

#### One-way Within-Subjects ANOVA | APA Style ANOVA Tables

```{r}

apa.ezANOVA.table(res,
                  correction="none",
                  table.title = "Repeated Measure ANOVA",
                  table.number=3)

```