---
title: "Analysis of Covariance Tutorials: One-way and Factorial Analyses"
author: "Professor Christopher J. Schmank"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r warning=FALSE,message=FALSE}
# Load in Required Libraries
library(psych)
library(tidyverse)
library(jmv)
library(ggpubr)
library(apaTables)
library(ez)
library(rstatix)
```

## ANCOVA | One-Way Research Context Prompt

**Research Context**:

 Researchers investigated the effect of exercises in reducing anxiety levels. Anxiety scores were measured from three groups of individuals practicing physical exercises at different levels (`low`, `moderate`, and `high`).

The anxiety score was measured `pre-` and 6-months `post-exercise` training programs. It is expected that any reduction in the anxiety by the exercises programs would also depend on the participant’s baseline level of anxiety.

In this analysis we use the `pre-test anxiety score` as the covariate and are interested in possible differences between `group` with respect to the `post-test anxiety score`.

## One-way ANCOVA | Data
```{r}
dat_anx <- read.csv("ancova1.csv")
```

## ANCOVA | Descriptive Stats (Baseline-test)

```{r}
dat_anx$groupF <- factor(dat_anx$group, 
                         levels = c("Low", "Moderate", "High"), 
                         labels = c("Low", "Moderate", "High"))

# Descriptive Stats
describeBy(dat_anx$pretest,   
           dat_anx$groupF,
           mat=TRUE)     

describeBy(dat_anx$posttest,  
           dat_anx$groupF, 
           mat=TRUE)      
```

## ANCOVA | EDA Visualiations

```{r}
# Histograms
ggplot(data = dat_anx,
       mapping = aes(x = pretest)) +
  geom_histogram(bins = 8, color = "coral") +
  labs(y = "Count", x = "Baseline Anxiety")+
  facet_wrap(dat_anx$groupF,ncol=1)

ggplot(data = dat_anx,
       mapping = aes(x = posttest)) +
  geom_histogram(bins = 8, color = "coral") +
  labs(y = "Count", x = "Post-Treatment Anxiety")+
  facet_wrap(dat_anx$groupF,ncol=1)

```

```{r}
# Boxplots
ggplot(data = dat_anx,
       mapping = aes(y = pretest, x = groupF)) +
  geom_boxplot() +
    labs(y = "Baseline Anxiety",
         x = "Exercise Group")

ggplot(data = dat_anx,
       mapping = aes(y = posttest, x = groupF)) +
  geom_boxplot() +
    labs(y = "Post-Treatment Anxiety",
         x = "Exercise Group")
```

```{r}
# Violin Plots

ggplot(data = dat_anx,
       aes(x = groupF, y = pretest, 
           fill = groupF))+
  geom_violin(show.legend=FALSE)+
  ylab("Baseline Anxiety")+
  xlab("Exercise Group")+
  scale_fill_manual(values=c("#D55e00","#56B4E9","#009E73"))

ggplot(data = dat_anx,
       aes(x = groupF, y = posttest, 
           fill = groupF))+
  geom_violin(show.legend=FALSE)+
  ylab("Post-treatment Anxiety")+
  xlab("Exercise Group")+
  scale_fill_manual(values=c("#D55e00","#56B4E9","#009E73"))
```

## ANCOVA | Omnibus Assumption Check 

```{r}
ancova(data = dat_anx,       
       dep = posttest,       
       factors = groupF,      
       covs = pretest,       
       effectSize = "eta",   
       homo = TRUE,          
       qq = TRUE,            
       norm = TRUE)          
```

## ANCOVA | Linearity Assumption Check
```{r}

ggscatter(dat_anx,                          
          x = "pretest",                    
          y = "posttest",                   
          color = "group",
          add = "reg.line",
          xlab = "Pre-test Scores",
          ylab = "Post-test Scores")+
  stat_regline_equation(    
    aes(label = paste(after_stat(eq.label), 
                      after_stat(rr.label), 
                      sep = "~~~"),         
        color = group))
```

## ANCOVA | Homogeneity of Regression Slopes Assumption Check
```{r}
dat_anx %>%
  anova_test(posttest ~ groupF*pretest)   

# Since p > .05 for the interaction we can assume homogeneity of regression slopes
```

## ANCOVA | Normality of Residuals

```{r}
lm<-lm(posttest~pretest + groupF, 
       dat_anx)

model.metrics <- augment(lm)       

shapiro_test(model.metrics$.resid)  

# Lack of statistical significant implies that normality of residuals has been achieved
```

## ANCOVA | Homogeneity of Variances
```{r}

model.metrics %>%                 
  levene_test(.resid ~ groupF)     

# Lack of statistical significance here implies that residual variances are homogeneous across between-subject groups
```

## ANCOVA | Conducting the ANCOVA

```{r}
options(digits = 3)
ancova(data = dat_anx,         
       dep = posttest,            
       factors = groupF,     
       covs = pretest,             
       effectSize = "eta",     
       postHoc = 'groupF',      
       postHocCorr = 'bonf',
       postHocES = 'd',     
       postHocEsCi = TRUE,  
       emMeans=list(
          list('groupF')),  
       emmTables=TRUE,        
       emmPlots=TRUE)         
``` 

## ANCOVA | Professional ANCOVA Visualization

```{r}

res.aov<- dat_anx %>% 
  anova_test(dv = posttest,
             between= group,
             covariate = pretest,
             effect.size = 'ges')

# Generates Marginal Means for Visualization
pwc <- dat_anx %>% 
  emmeans_test(posttest ~ group,
               covariate = pretest,
               p.adjust.method = "bonferroni") %>%
  add_xy_position(x = "group")                                   

pwc
```

## ANCOVA | Professional ANCOVA Visualization

```{r}
get_emmeans(pwc)
ggline(get_emmeans(pwc),
       x = "group",
       y = "emmean") +                           
  geom_errorbar(aes(ymin = conf.low,             
                    ymax = conf.high),
                width = 0.1) +                   
  stat_pvalue_manual(pwc,                        
                     hide.ns = TRUE,             
                     tip.length = FALSE) +       
  ylab("Mean Post-Test Anxiety") +               
  xlab("Group") +                                
  labs(caption = get_pwc_label(pwc)                 
  )

```


## ANCOVA | Factorial Research Context Prompt

Your local sports psychologist and stress research enthusiast found out you were learning various quantitative statistical methods and reached out to offer you a consulting/analytics opportunity. They gathered data from 60 participants to attempt to evaluate the effect of a new protein treatment and exercise on experienced stress levels in a group of older adults. The data set (`dat_stress`) contains five variables:

- `id` --- Participant identification number  
- `score` --- Stress score collected from each participant
- `treatment` --- Treatment variable with two levels (`yes` --- Received protein treatment; `no` --- Received placebo treatment)
- `exercise` --- Level of exercise employed by participants (`low`, `moderate`, `high`)
- `age` --- Age of participant when data was collected

The goal for this final research question was: Does adjusting for the age of participants impact how stress scores fluctuate across treatment and exercise variables?

## Factorial ANCOVA | Data
```{r}
dat_stress <- read.csv("ancova2.csv")
```

## Factorial ANCOVA | Descriptive Stats (Baseline-test)

```{r}

# Descriptive Stats
dat_stress$Treatment <- factor(dat_stress$treatment,
                                levels = c("no", "yes"),
                                labels = c("No Treatment", "Protein Treatment"))

dat_stress$Exercise <- factor(dat_stress$exercise, 
                               levels = c("low", "moderate", "high"),
                               labels = c("Low", "Moderate", "High"))

describeBy(dat_stress$score,   
           list(dat_stress$Treatment, dat_stress$Exercise),
           mat=TRUE) 

```

## Factorial ANCOVA | EDA Visualizations

```{r}

# Histogram
ggplot(data = dat_stress,
       mapping = aes(x = score)) +
  geom_histogram(bins = 8, color = "goldenrod2") +
  labs(y = "Count", x = "Stress Score")+
  facet_wrap(~dat_stress$Treatment + dat_stress$Exercise)

# Boxplots
ggplot(data = dat_stress,
       mapping = aes(y = score, x = Treatment)) +
  geom_boxplot() +
    labs(y = "Stress Score")

ggplot(data = dat_stress,
       mapping = aes(y = score, x = Exercise)) +
  geom_boxplot() +
    labs(y = "Stress Score")

# Violin Plot
ggplot(data = dat_stress,
       aes(x = Exercise, y = score, 
           fill = Treatment))+
  geom_violin(show.legend=TRUE)+
  ylab("Stress Score")+
  xlab("Exercise Type")+
  scale_fill_manual(values=c("#D55e00","#56B4E9","#009E73"))

```


## Factorial ANCOVA | Assumption Check: Omnibus and Grouped Level 

```{r}

# Check Assumptions using techniques described in class

ancova(data = dat_stress,       
       dep = score,             
       factors = c(Treatment,Exercise),      
       covs = age,
       effectSize = "partEta",
       homo = TRUE,          
       qq = TRUE,
       norm = TRUE)

# Grouped Assumption Checks

# Linear relationship based on age and IVs
ggscatter(dat_stress,  
          x = "age",                         
          y = "score",                       
          facet.by = c("Exercise","Treatment")) +
  stat_smooth(method = "loess")

# Homogeneity of Regression slopes

dat_stress %>%
  anova_test(score ~ age + Treatment + Exercise + Treatment*Exercise + age*Treatment + age*Exercise + age*Exercise*Treatment)

# Normality assessments
lm<-lm(score~age + Exercise*Treatment, dat_stress)      
model.metrics <- augment(lm)       
shapiro_test(model.metrics$.resid)

# Homogeneity of variance assessments
model.metrics %>%  
  levene_test(.resid ~ Exercise*Treatment)
```

## Factorial ANCOVA | Conducting the ANCOVA

```{r}
options(digits=3)
ancova(data = dat_stress,
       dep = score,               
       factors = c(Treatment,Exercise),
       covs = age,              
       effectSize = "partEta",
       postHoc = ~Exercise + Treatment:Exercise,     
       postHocCorr = 'bonf',  
       postHocES = 'd',    
       postHocEsCi = TRUE, 
       emMeans= ~ Exercise + Treatment + Exercise:Treatment,
       emmTables=TRUE,
       emmPlots=TRUE) 

# Assessment of Simple Main Effects
dat_stress %>% 
  group_by(Exercise) %>% 
  anova_test(score ~ age + Treatment)

dat_stress %>% 
  group_by(Treatment) %>% 
  anova_test(score ~ age + Exercise)
``` 

## Factorial ANCOVA | Professional ANCOVA Visualization

```{r}
# Pairwise Comparisons
pwc <- dat_stress %>% 
  group_by(Exercise) %>% 
  emmeans_test(score ~ Treatment,
               covariate = age,
               p.adjust.method = "bonferroni")

pwc %>% filter(Exercise == "High") 

pwc2 <- dat_stress %>% 
  group_by(Treatment) %>% 
  emmeans_test(score ~ Exercise,
               covariate = age,
               p.adjust.method = "bonferroni")

pwc2 %>% filter(Treatment == "Protein Treatment")

```

```{r}

lp <- ggline(
  get_emmeans(pwc),
  x = "Exercise",
  y = "emmean",
  color = "Treatment",
  palette = "grey") +
  geom_errorbar(
    aes(ymin = conf.low,
        ymax = conf.high),
    width = 0.1) 
```

```{r}
pwc <- pwc %>% add_xy_position(x = "Exercise",
                  fun = "mean_se",
                  step.increase = 0.2)

pwc.filtered <- pwc %>% filter(Exercise == "High")


lp + stat_pvalue_manual(pwc,
                     hide.ns = TRUE,
                     tip.length = 0) +
  labs(
    y = "Estimated Means",
    x = "Exercise Condition",
    color = "Treatment",
    caption = get_pwc_label(pwc)
  )
```

```{r}
pwc2 <- pwc2 %>% add_xy_position(x = "Exercise",
                  fun = "mean_se")

pwc2.filtered <- pwc2 %>% filter(Treatment == "Protein Treatment")


lp + stat_pvalue_manual(pwc2,
                     hide.ns = TRUE,
                     tip.length = 0) +
  labs(
    y = "Estimated Means",
    x = "Exercise Condition",
    color = "Treatment",
    caption = get_pwc_label(pwc)
  )
```