---
title: "One-Way Within-Subjects ANOVA Demo"
author: "Christopher J. Schmank, PhD"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### One-way Within-Subjects ANOVA | Research Context Prompt

A local stress researcher is interested in the relationship between psychosocial stress levels and processing speed. After conducting a brief literature review they have hypothesized that regulating feelings of stress impairs concurrent performance of processing speed.

The researcher conducted a study where psychosocial stress was manipulated into three conditions: low stress (**Calm**), mild stress (**Mild**), and strong stress (**Strong**). Money was tight and recruitment was difficult, so they implemented a within-subjects, repeated measures design. Data was collected from 20 participants that consented to participate. The dependent measure in this task was processing speed ability measured in 10-second intervals (i.e., 1.3 = 13 seconds!!)

#### One-way Within-Subjects ANOVA | Data

```{r message = FALSE, warning = FALSE}

library(psych)
library(tidyverse)
library(jmv)
library(ggpubr)
library(ggrain)
library(apaTables)
library(ez)
library(rstatix)

dat_stress <- read.csv("RM.ANOVA.Demo.csv")
dim(dat_stress)
head(dat_stress)

```

#### One-way Within-Subjects ANOVA | Descriptive Stats

```{r}

# Descriptive Stats
describe(dat_stress[-1])

```

### One-way Within-Subjects ANOVA | Visualizations

#### First we need to make the wide data into a long format

```{r}
# Create Long Data Frame using tidyr::gather()

dat_stressL<- dat_stress %>% 
  gather(key="Stress_Condition",
         value="Speed",
         Calm,
         Mild,
         Strong,
         factor_key = TRUE)

dat_stressL$ID <- factor(dat_stressL$ID)

# OR you can use the tidyr::pivot_longer() function workflow

dat_stressL2 <- dat_stress %>% 
  pivot_longer(cols = -ID,
               names_to = "Stress_Condition",
               values_to = "Speed")

dat_stressL2$Stress_Condition<-factor(dat_stressL2$Stress_Condition,labels = c("Calm","Mild","Strong"))

dat_stressL2$ID <- factor(dat_stressL2$ID)
```

```{r}
# Boxplot Visualization
ggplot(data = dat_stressL,
       mapping = aes(y = Speed, x = Stress_Condition)) +
  geom_boxplot() +
  ylab("Processing Speed (10-second Intervals)") +
  xlab("Stress Condition")

# Raincloud Plot Visualization
ggplot(data = dat_stressL,
       aes(x = Stress_Condition, y = Speed, 
           fill = Stress_Condition))+
  geom_rain(show.legend=FALSE) +
  scale_colour_grey(aesthetics = "fill",
                    start = .6,
                    end = 1) +
  coord_flip() +
  ylab("Processing Speed (10-Second Intervals)")+
  xlab("Stress Condition")+
  theme_classic()

```

#### One-way Within-Subjects ANOVA | Sphericity Violation Check

```{r}
# JMV Code does NOT require Long data frame object

anovaRM(data=dat_stress,
        rm = list(
          list(label = 'Stress Condition',
               levels = list('Calm','Mild','Strong'))),
        rmCells = list(
          list(measure = 'Calm', cell = 'Calm'),
          list(measure = 'Mild', cell = 'Mild'),
          list(measure = 'Strong', cell = 'Strong')),
        rmTerms= list('Stress Condition'),
        spherTests = TRUE)

```

#### One-way Within-Subjects ANOVA | Normality Assumption Check

```{r}
# ggpubr::ggqqplot() requires long data frame

dat_stressL %>% 
  group_by(Stress_Condition) %>% 
  shapiro_test(Speed)

ggqqplot(dat_stressL,
         'Speed',
         facet.by = "Stress_Condition")

```


#### One-way Within-Subjects ANOVA | Conducting the ANOVA 

```{r}
options(digits = 3)
anovaRM(data=dat_stress,
        rm = list(
          list(label = 'Stress',levels = list('Calm','Mild','Strong'))),
        rmCells = list(
          list(measure = 'Calm', cell = 'Calm'),
          list(measure = 'Mild', cell = 'Mild'),
          list(measure = 'Strong', cell = 'Strong')),
        rmTerms= list('Stress'),
        effectSize = 'eta',
        postHoc='Stress',
        postHocCorr = list("bonf"),
        emMeans=list(
          list('Stress')),
        emmTables=TRUE,
        emmPlots=FALSE,
        spherTests = TRUE,
        spherCorr = list('none', 'GG'))

```
#### One-way Within-Subjects ANOVA | ANOVA Output: Manual Post-Hoc Tests Code

```{r}

pwc <- dat_stressL %>%
  pairwise_t_test(Speed~Stress_Condition,
                  paired=TRUE, 
                  p.adjust.method = "bonferroni") %>%
  add_xy_position(x = "Stress_Condition")

```

#### One-way Within-Subjects ANOVA | Professional ANOVA Visualization Code

```{r}

ggboxplot(dat_stressL, 
          x = "Stress_Condition", 
          y = "Speed") +
  stat_pvalue_manual(pwc,
                     #hide.ns = TRUE, 
                     tip.length = 0) +
  labs(y = "Processing Speed (10-second Intervals)", x = "Stress Level")

```

#### One-way Within-Subjects ANOVA | APA Style ANOVA Tables

```{r}
model <- ezANOVA(dat_stressL,
                 dv=.(Speed),
                 wid=.(ID),
                 within=.(Stress_Condition),
                 type = 3,
                 detailed=TRUE)

apa.ezANOVA.table(model,
                  correction="none",
                  table.title = "Repeated Measure ANOVA",
                  table.number=3)

```