---
title: "One-Way Between-Subjects ANOVA Demo"
author: "Christopher J. Schmank, PhD"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### One-way Between-Subjects ANOVA | Research Context Prompt

A developmental psychologist is looking for a Grad student who can help them with their research project looking into adolescent self-esteem and its relationship with parenting style. There were three parenting styles: (1) `authoritative` (high expectations with support), (2) `authoritarian` (high expectations with little support), and (3) `permissive` (low expectations with support).

Data on 90 teenagers self-esteem levels and the parenting style they experienced (30 per style) has been collected. The research question of interest is: Does parenting style differences impact teenage self-esteem?

#### One-way Between-Subjects ANOVA | Data

```{r message = FALSE, warning = FALSE}
library(psych)
library(tidyverse)
library(ggrain)
library(jmv)
library(ggpubr)
library(apaTables)
library(rstatix)

dat_A <- read.csv("One-way ANOVA.A.csv")
dim(dat_A)
head(dat_A)

```

#### One-way Between-Subjects ANOVA | Descriptive Stats

```{r}

# Descriptive Stats
describe(dat_A[,2:3])

```

#### One-way Between-Subjects ANOVA | Grouped Descriptive Stats

```{r}

dat_A$Style<-factor(dat_A$Style, 
                    labels = c("Authoritative","Authoritarian","Permissive"))

# Descriptive Stats
describeBy(dat_A$Self_esteem,
           dat_A$Style, 
           mat=TRUE)

```

### One-way Between-Subjects ANOVA | Univariate Visualizations

#### One-way Between-Subjects ANOVA | Grouped Boxplot

```{r}

ggplot(data = dat_A,
       mapping = aes(y = Self_esteem, x =Style)) +
  geom_boxplot() +
    labs(y = "Parenting Style")

```

#### One-way Between-Subjects ANOVA | Raincloud Plot

```{r}
ggplot(data = dat_A,
       aes(x = Style, y = Self_esteem, 
           fill = Style))+
  geom_rain(show.legend=FALSE) +
  scale_colour_grey(aesthetics = "fill",
                    start = .6,
                    end = 1) +
  coord_flip() +
  ylab("Self-Esteem")+
  xlab("Parenting Style")+
  theme_classic()


ggplot(data = dat_A,
       aes(x = Style, y = Self_esteem, 
           fill = Style))+
  geom_rain(show.legend=FALSE) +
  coord_flip() +
  ylab("Self-Esteem")+
  xlab("Parenting Style")+
  scale_fill_manual(values=c("#D55e00","#56B4E9","#009E73"))+
  theme_classic()
```

#### One-way Between-Subjects ANOVA | Omnibus Assumption Checks Code

```{r}

# Omnibus Assumption Checking
ANOVA(data = dat_A, 
      dep = 'Self_esteem', 
      factors = c('Style'),
      homo = TRUE,
      norm = TRUE,
      qq = TRUE)

```

#### One-way Between-Subjects ANOVA | Group Level Assumption Checks 

```{r}

dat_A %>% 
  group_by(Style) %>% 
  shapiro_test(Self_esteem)

```

#### One-way Between-Subjects ANOVA | Group Level Assumption Checks

```{r}

ggqqplot(dat_A,
         'Self_esteem',
         facet.by = "Style")

```

#### One-way Between-Subjects ANOVA | Rechecking Homogen. of Variance using `rstatix`

```{r}

dat_A %>% 
  levene_test(Self_esteem ~ as.factor(Style), center = mean)

```

#### One-way Between-Subjects ANOVA | Conducting the ANOVA Code

```{r}
options(digits = 3)
ANOVA(data = dat_A, 
      dep = 'Self_esteem', 
      factors = c('Style'), 
      effectSize = 'eta', 
      postHoc = 'Style', 
      postHocCorr = 'bonf',
      postHocES = 'd',
      postHocEsCi = TRUE,
      emMeans = list(
       list('Style')),
      emmPlots = FALSE,
      emmTables = TRUE)

```

### One-way Between-Subjects ANOVA | Professional ANOVA Visualization Code

#### One-way Between-Subjects ANOVA | Boxplot Visualization w/ Significance Indicators Code

```{r}

# Pairwise Post-Hoc Tests (Bonferroni Correction)
pwc <- dat_A %>% 
  pairwise_t_test(Self_esteem ~ Style, p.adjust.method = "bonferroni") %>% 
  add_xy_position(x = "Style")

# Visualization: box plots with p-values

ggboxplot(dat_A, 
          x = "Style", 
          y = "Self_esteem", 
          #notch = TRUE, 
          outlier.shape = NA) +
  stat_pvalue_manual(pwc, 
                     hide.ns = TRUE, 
                     tip.length = 0) +
  labs(y = "Self Esteem")

```

#### One-way Between-Subjects ANOVA | APA Style ANOVA Tables

```{r}
lm<-lm(Self_esteem ~ Style, dat_A)
apa.aov.table(lm,
              table.number=1,
              conf.level=.95,
              type = 2)

apa.1way.table(Style,
               Self_esteem,
               dat_A,
               table.number=2,
               show.conf.interval = TRUE)
```
