---
title: "Three-way Factorial Analysis of Variance Tutorial"
author: "Professor Christopher J. Schmank"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Research Context**:

A Psycho-pharmacology firm has hired you to be their data analyst and provided you with a data set that consists of two demographic variables, a treatment variable, and current patient cholesterol ratings. The data set (`dat3`) contains five variables:

- `id` --- Participant identification number  
- `gender` --- Demographic variable with two levels (`male` and `female`)  
- `risk` --- Demographic variable with two levels (`low` --- Low risk for heart attack and `high` --- High risk for heart attack)
- `drug` --- Treatment variable with three levels (`A` --- Placebo Treatment; `B` --- Niacin Treatment; and `C` --- Plant Sterol Treatment)
- `cholesterol` --- Amount of cholesterol detected 

The goal for this first research question was: Does participants gender, risk, and treatment interactively impact cholesterol? 

```{r message = FALSE, warning = FALSE}
library(psych)
library(tidyverse)
library(jmv)
library(ggpubr)
library(apaTables)
library(ez)
library(rstatix)

dat <- read.csv("anova3.csv")
dat2 <- read.csv("anova3.csv")

dat$gender<-factor(dat$gender, labels = c("Female","Male"))
dat$risk<-factor(dat$risk, labels = c("High", "Low"))
dat$drug <- factor(dat$drug, labels = c("Placebo", "Niacin", "Plant Sterol"))
```

#### Factorial ANOVA | Descriptive Stats

```{r}
# Descriptive Stats

describeBy(dat$cholesterol, 
           list(dat$gender,dat$risk,dat$drug),       
           mat=TRUE) 

```

#### Factorial ANOVA | Histogram

```{r }
ggplot(data = dat,
       mapping = aes(x = cholesterol)) +
  geom_histogram(bins = 8, color = "maroon4") +
  labs(y = "Count", x = "Cholesterol")+
  facet_wrap(~dat$gender+dat$risk+dat$drug, ncol =3)
```

#### Factorial ANOVA | Boxplot

```{r}

ggplot(data = dat,
       mapping = aes(y = cholesterol, x = gender)) +
  geom_boxplot() +
    labs(y = "Cholesterol",
         x = "Gender")
```


#### Factorial ANOVA | Boxplot

```{r }

ggplot(data = dat,
       mapping = aes(y = cholesterol, x = drug, group = risk)) +
  geom_boxplot() +
    labs(y = "Cholesterol",
         x = "Treatment")+
  facet_wrap(c("risk", "gender"))

```


#### Factorial ANOVA | ANOVA() Omnibus Assumption Checks

```{r}

# Omnibus Assumption Checking
ANOVA(data = dat, 
      dep = 'cholesterol',             
      factors = c('gender','risk', 'drug'),
      homo = TRUE,                      
      norm = TRUE,                      
      qq = TRUE)      

```

#### Factorial ANOVA | Group Level Assumption Checks

```{r}
# Normality Assessment
model  <- lm(cholesterol ~ gender*risk*drug, data = dat)
# Create a QQ plot of residuals
ggqqplot(residuals(model))
# Compute Shapiro-Wilk test of normality
shapiro_test(residuals(model))

# Normality by groups
dat %>%
  group_by(gender, risk, drug) %>%
  shapiro_test(cholesterol)

ggqqplot(dat, 
         "cholesterol", 
         ggtheme = theme_bw()) +
  facet_grid(gender + risk ~ drug, 
             labeller = "label_both")

dat %>%   
  levene_test(cholesterol ~ gender*risk*drug, 
              center = mean) 

```

#### Factorial ANOVA | Conducting the ANOVA

```{r}
# Between-Subject Factorial ANOVA (jmv)
ANOVA(data = dat,                    
      dep = 'cholesterol',               
      factors = list('gender','risk','drug'),  
      effectSize = 'partEta',             
      postHoc = ~drug + drug:gender + gender:risk:drug, 
      postHocCorr = 'bonf',
      postHocES = 'd',   
      postHocEsCi = TRUE,
      emMeans = ~gender + risk + drug + drug:risk + drug:gender + drug:gender:risk,         
      emmPlots = TRUE,
      emmPlotData = TRUE,
      emmTables = TRUE)   

# Simple two-way interaction effect check Risk and Drug across Gender
model <- lm(cholesterol~drug*gender*risk, dat)

sie.check <- dat %>%                 
  group_by(risk) %>%                  
  anova_test(dv = cholesterol,         
             between = c(gender,drug),
             error = model)%>%  
  get_anova_table() %>%                 
  adjust_pvalue(method = "bonferroni")  
sie.check

# Simple simple main effects
ssme.check <- dat %>%
  group_by(gender,risk) %>% 
  anova_test(dv = cholesterol,         
             between = drug,
             error = model)%>%
  get_anova_table() %>%                 
  adjust_pvalue(method = "bonferroni")  
ssme.check
```

#### Factorial ANOVA Visualization

```{r}
pwc <- dat %>%
  group_by(gender, risk) %>%
  emmeans_test(cholesterol ~ drug, p.adjust.method = "bonferroni")
pwc

```
```{r}
pwc %>% filter(gender == "Male", risk == "High")

get_emmeans(pwc) %>% filter(gender == "Male", risk == "High")

```

```{r}
pwc <- pwc %>% add_xy_position(x = "drug")
pwc.filtered <- pwc %>% filter(gender == "Male", risk == "High")

ggboxplot(dat,
          x = "drug", 
          y = "cholesterol",
          color = "risk", 
          palette = "grey", 
          facet.by = "gender") +
  stat_pvalue_manual(pwc.filtered, 
                     color = "risk", 
                     hide.ns = TRUE,
                     tip.length = 0, 
                     step.increase = 0.1, 
                     step.group.by = "gender") +
  labs(x = "Drug Treatment Provided",
       y = "Cholesterol",
       color = "Risk of Heartattack",
       caption = get_pwc_label(pwc))


```

