---
title: "Covariance, Correlation, and Regression Tutorial"
author: "Professor Christopher J. Schmank"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### **Load Libraries**

```{r echo = TRUE, warning=FALSE, message=FALSE}
library(tidyverse)
library(mice)
library(psych)
library(jmv)
library(apaTables)
```

#### **Variance and Covariance Assessment: Complete Data Examples**

```{r echo=TRUE}
var(attitude)

cov(attitude)

```

- These functions will work interchangeably by default IF AND ONLY IF there is no missing data!

#### **Variance and Covariance Assessment: Incomplete Data Examples**

```{r echo=TRUE}
var(nhanes)

cov(nhanes) # By default this function sets `use = "everything"` 

```

- When missing data is present and default `var()` and `cov()` are used they result in same variance-covariance matrices---unusable in this form!!

```{r}
var(nhanes, na.rm=TRUE) #All cases with NA removed

cov(nhanes, use = "complete.obs") #All cases with NA removed

cov(nhanes, use = "pairwise.complete.obs") #Pairwise deletion --- different N's

```

#### **Calculating Correlation Matrix**

Correlations can be generated with several different functions in base R but get familiar with the `psych` and `JMV` package function calls for correlation matrices

**1. If all you want is a correlation matrix without any additional information**: Use `cor()`

```{r}
cor(attitude)
cor(attitude[1:4]) #Can use subsetting to only pull certain variables into matrix
cor(nhanes) #If missing data is present you must remove it or set `use =` argument
```

**2. If you want to get correlation on ONLY 2 variables at once**: Use `cor.test()`

```{r}
cor.test(~rating + learning, data = attitude)
cor.test(x=attitude$rating, y=attitude$learning)
```

- Provides `*t*-test`, `degrees of freedom`, and `*p*-value` to assess if correlation is statistically significantly different from zero

- Provides `correlation coefficient` and `95% confidence interval`

- Provides type of correlation coefficient calculated (i.e., `Pearson's product-moment correlation`)

**3. For more detail and details needed for publication**: Use `corr.test()` or `corrMatrix`

```{r}
corr.test(attitude)

corrMatrix(attitude,
           vars = vars(rating, complaints, privileges,learning))
```

**4. If you need to save a correlation matrix for later use**: Use `corr.test()` or `cor()`

```{r}
cor_dat <- corr.test(attitude[1:4])

cor_dat$r # Calls for ONLY the correlation matrix portion of corr.test() output

cor_dat_r <- cor_dat$r # May also want to make an R object for the correlations

```

```{r}
cor_dat2 <- cor(attitude) # This code creates a similar object as `cor_dat_r` above
cor_dat2
```

#### **Creating Correlation Table Document**

The `apaTables` package can be used to create quick correlation matrices and linear regression tables---these can be helpful but are VERY difficult to reformat or make changes to.

Make sure that you know how to create these tables using other programs like Excel and Word as these allow for much more customization!

```{r}
apa.cor.table(data = attitude[1:4],
              filename = "attitude-correlation-matrix.doc",
              table.number = 1)
```

#### **Writing Up Correlation**

Can be achieved with three statements:

1. Was correlation statistically significantly different from zero; what was the magnitude and direction (adding confidence intervals is a good practice)?
1. How does the magnitude compare to correlation cutoff values?
1. Does result align with predictions/hypotheses?

The correlation between `attitude rating` and `being too critical ratings` was not statistically significantly different from zero, *r*(28) = .16, 95% CI [-.22, .16], *p* = .41. According to criteria provided by Cohen (1988), the size (i.e., magnitude) of this correlation fell above a `weak` correlation, but well below what would be considered `moderate.` The sign of the correlation was `positive`, however the lack of statistical significance implies that this direction is not stable within this sample of data.

#### **Conducting and Writing Up Linear Regression**

```{r}
# Easy coding method requires linear model 
# Followed by `anova()` and `summary()` calls for description of model
# Output is less than ideal but coding is efficient
lm <- lm(rating ~ privileges, data = attitude)
anova(lm)
summary(lm)

# JMV coding method requires several argument calls as shown below
# Output is clean but coding gets more difficult with more predictors
linReg(attitude, 
       dep = rating, 
       covs = ("privileges"), 
       blocks = list(
         list("privileges")),
       modelTest = TRUE,
       stdEst = TRUE,
       ciStdEst = TRUE)

```

The linear regression predicting `attitude ratings` from `ratings towards allowing special privileges` indicated that `allowing special privileges` accounted for approximately 18% of the variance in `attitude ratings`, *F*(1,28) = 6.21, *p* = .02, *R^2^* = .182. The unstandardized and standardized regression coefficients were statistically significantly different from zero, *b* = 0.424, $\beta$ = .426 (95% CI [.08, .78]), *t*(29) = 2.49, *p* = .02.The value of these coefficients respectively demonstrate: (1) for every one-unit increase in `allowing special privileges` there was an increase of 0.424 units in `attitude ratings` and (2) for every one standard deviation unit increase in `allowing special privileges` there was an increase of 0.426 standard deviation units in `attitude ratings`.

#### **Creating Regression Table Document**

```{r warning=FALSE, message=FALSE}
# Generate linear model using
lm <- lm(rating~privileges, data = attitude)

# Create Word Document with APA style Regression Table

apa.reg.table(lm,
              filename = "regression_table.doc",
              table.number = 2)

apa.reg.boot.table(lm,
                   filename = "regression_boot.doc",
                   table.number = 2,
                   number.samples = 1000)
```

