---
title: 'Multiple Regression: Moderation and Mediation Tutorial'
author: "Christopher J. Schmank, PhD"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load in required libraries

```{r message=FALSE,warning=FALSE}
library(jmv)
library(rockchalk)
library(psych)
library(multilevel)
library(tidyverse)
library(interactions)
```


#### **Load data set into R-Environment**
```{r}

dat<-read.csv("DemoMod.csv")
dat2<-read.csv("DemoMed.csv")

```


### Moderation Tutorial Markdown

We are going to use `dat` for a tutorial of moderation analyses. This data set consists of performance evaluation review data (*perf_eval*) for employees of a Marketing Firm. Additionally, we have data for the employee's original interview rating (*interview*), their age at hire (*age*), dichotomous coded sex (*gender*: 0 = Males; 1 = Females), and polytomous race (*race*: 1 = Caucasian, 2 = African-American, 3 = Asian).

**Applied prompt**: Our boss wants to know if the association between performance evaluation and interview ratings is moderated by the demographic variables of age and gender. We will treat performance evaluation as the criterion variable for this  tutorial.

#### **Recode Predictors and Moderators (Centering and Dummy Coding)**

```{r}
# Center all predictors/moderators that are continuous
dat$interviewC <- dat$interview - mean(dat$interview) 
dat$ageC <- dat$age - mean(dat$age) 

# Dummy code all categorical variables NOT already coded 0 and 1 
# These will NOT be used in the analyses, but you could use them to expand your understanding
AA_dc <- dummy.code(dat$race, group = 2)
As_dc <- dummy.code(dat$race, group = 3)
dat <- data.frame(dat, AA_dc, As_dc)

# Create our interaction term(s)
dat$int_gender <- dat$interviewC*dat$gender
dat$int_age <- dat$interviewC*dat$ageC
```

#### **Calculate Descriptive Statistics and Correlation Matrix**

```{r}
dat$genderF <- factor(dat$gender,
                     levels = c(0,1), 
                     labels = c("Women","Men"))

describe(dat[2:4])

describeBy(dat[2:4],
           group = dat$genderF,
           mat=TRUE)

corrMatrix(dat[,c(2:5)],
           plots = TRUE,
           plotDens = TRUE,
           plotStats = TRUE)$plot

```

#### **Moderation | `linReg()` Technique** Interview By Gender Moderation


```{r}

linReg(data = dat, 
       dep = 'perf_eval', 
       covs = c('interviewC', 'gender','int_gender'),
       blocks = list(                                  
         list('interviewC', 'gender'),                 
         list('int_gender')),                         
       ci = TRUE,
       modelTest = TRUE,
       r2Adj = TRUE,
       stdEst = TRUE,
       ciStdEst = TRUE)
```


#### **Moderation | `lm()` Technique** Interview By Gender Moderation

```{r}
# Interview x Gender Moderation Models

# Step 1: Unstandardized coefficients 
no_mod1 <- lm(perf_eval ~ interviewC + gender, data = dat)
# Step 2: Unstandardized coefficients 
mod1 <- lm(perf_eval ~ interviewC*gender, data = dat)

# Step 1: Standardized coefficients
no_mod1Z <- lm(scale(perf_eval) ~ scale(interviewC) + scale(gender), data = dat)
# Step 2: Standardized coefficients (!!YOU MUST WRITE OUT ENTIRE MODEL UNLIKE mod1 ABOVE!!)
mod1Z <- lm(scale(perf_eval) ~ scale(interviewC) + scale(gender) + scale(interviewC*gender), data = dat)

# Direct model comparisons
anova(no_mod1, mod1)
# anova(no_mod1Z, mod1Z) ## Can uncomment to assess standardized model (same output as unstandardized)

# Model Summary
summary(mod1Z)
# summary(no_mod1Z)
```

```{r}
# Visualization of Moderation Plots
plotSlopes(mod1,                            
           plotx = "interviewC",                
           modx = "gender",                     
           main = "Gender Moderates Interview Rating Impact on Performance Eval",
           xlab = "Interview Score (Centered)", 
           ylab = "Performance Evaluation",
           modxVals = "std.dev.",
           legendArgs = list(title = "Gender"))  
```

```{r}
johnson_neyman(mod1,
               pred = "interviewC",
               modx = "gender",
               control.fdr = TRUE)
```


#### **Moderation | `linReg()` Technique** Interview By Age Moderation

```{r}
linReg(data = dat, 
       dep = 'perf_eval', 
       covs = c('interviewC', 'ageC','int_age'),
       blocks = list(                                  
         list('interviewC', 'ageC'),                 
         list('int_age')),                         
       ci = TRUE,
       modelTest = TRUE,
       r2Adj = TRUE,
       stdEst = TRUE,
       ciStdEst = TRUE)

```

#### **Moderation | `lm()` Technique** Interview By Gender Moderation

```{r}
#Interview x Age Moderation Models
no_mod2 <- lm(perf_eval ~ interviewC + ageC, data = dat)
no_mod2Z <- lm(scale(perf_eval) ~ scale(interviewC) + scale(ageC), data = dat)

mod2 <- lm(perf_eval ~ interviewC*ageC, data = dat)
mod2Z <- lm(scale(perf_eval) ~ scale(interviewC*ageC), data = dat)

# Direct model comparisons
anova(no_mod2, mod2)
# anova(no_mod2Z, mod2Z) ## Can uncomment to assess standardized model (same output as unstandardized)

# Model Summary
summary(mod2Z)
# summary(no_mod2Z)

```

```{r}
# Visualization of Moderation Plots
plotSlopes(mod2,                            
           plotx = "interviewC",                
           modx = "ageC",                     
           main = "Age Moderates Interview Rating Impact on Performance Eval",
           xlab = "Interview Score (Centered)", 
           ylab = "Performance Evaluation",
           modxVals = "std.dev.",
           n = 5,
           legendArgs = list(title = "Mod: Age (Centered)")) 
```

```{r}
johnson_neyman(mod2,
               pred = "interviewC",
               modx = "ageC",
               control.fdr = TRUE)
```

---

### Mediation Tutorial Markdown

We are going to use `dat2` for a tutorial on mediation analyses. This data set consists of customer product review data. A product was reviewed by customers based on how much they enjoy the product (*Enjoy*), the quality of the product (*Qual*), how excited the customer was to purchase the item (*Excite*), how happy the purchase made the customer (*Happy*), and how much intent went into the purchase (*Intent*).

**Applied prompt**: The product manager wants to know how enjoyment of the product impacts quality ratings and whether the association between these variables can be accounted for by the intent of purchasing the product. We will treat quality as the criterion variable for this tutorial, with enjoyment and intent as predictor and mediator, respectively.

#### **Calculate Descriptive Statistics and Correlation Matrix**

```{r}

describe(dat2[2:6])

corrMatrix(dat2[2:6],
           plots = TRUE,
           plotDens = TRUE,
           plotStats = TRUE)$plot

```

#### **Mediation | `linReg()` Technique**

```{r}
#Indirect Path (a): M ~ X 

linReg(data=dat2,
       dep=Intent,
       covs=c(Enjoy),
       blocks=list(
         list('Enjoy')),
       ci = TRUE,
       modelTest = TRUE,
       r2Adj = TRUE,
       stdEst = TRUE,
       ciStdEst = TRUE)
```

```{r eval=FALSE,echo=TRUE}
# Step 1: Y ~ X (Path c) <br>
# Step 2: Y ~ X + Med (Paths b and c')
linReg(data=dat2,
       dep=Qual,
       covs=c(Enjoy,Intent),
       blocks=list(
         list('Enjoy'),
         list('Intent')),
       ci = TRUE,
       modelTest = TRUE,
       r2Adj = TRUE,
       stdEst = TRUE,
       ciStdEst = TRUE)
```

#### **Mediation | `lm()` Technique**

```{r}
# Standardized Direct Path (c)
lm1Z <- lm(scale(Qual)~scale(Enjoy), dat2)
lm1 <- lm(Qual ~ Enjoy, dat2)

# Standardized Indirect Path (a)
lm2Z <- lm(scale(Intent) ~ scale(Enjoy), dat2)
lm2 <- lm(Intent ~ Enjoy, dat2)

# Standardized Direct Path (c') and Indirect Path (b)
lm3Z <- lm(scale(Qual) ~ scale(Enjoy) + scale(Intent), dat2)
lm3 <- lm(Qual ~ Enjoy + Intent, dat2)

```

```{r}
# Model Comparison for Mediation Effect
anova(lm1Z, lm3Z)
```

```{r}
# Model Statistic Summaries
summary(lm1Z)
summary(lm2Z)
summary(lm3Z)
```

```{r}
# Conducting Sobel's Indirect Effect Test
sobel(pred=dat2$Enjoy, 
      med=dat2$Intent,
      out=dat2$Qual)
```

#### **Visualization of Mediation Model**
```{r}
mediate(Qual ~ Enjoy + (Intent), 
        data = dat2,
        std = TRUE,
        main = "Predicting Quality from Enjoyment Mediated by Intention"
        )$plot
```

