---
title: 'Multiple Regression Diagnostics and Simultaneous Multiple Regression Tutorial'
author: "Professor Christopher J. Schmank"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Multiple Regression Diagnostics Demonstration**

#### **Applied Prompt**:

A local resident recently joined their a newly developed charitable giving program and is interested in finding out more about `prosocial behavior` among group members. Without knowing much about the literature, they developed a short survey to give to a nearby school district to find out.

    **Research Question**: What influences the amount of money given to a program for community outreach. 

Ultimately, they must report their findings to their program coordinator and a statistician, who is also on the board.

#### **Variables in Data**:

`ID` --- identifying number for each observation

`Belief` --- belief that charitable giving has a positive effect on a scale of 1-10

`Need` --- rating of perceived amount of need required in the community on a scale of 1-10

`Interest` --- rating of level of interest in the project on a scale of 1-10

`Happy` --- rating of happiness felt when making donations on a scale of 1-10

`Amount` --- amount given from 0 - 10 dollars

#### **For all variables**: 

A value of `99` implies that a response was missing for that variable and case

#### **Load in Libraries**

```{r echo = TRUE, warning=FALSE, message=FALSE}
library(mice)
library(MVN)
library(lmtest)
library(apaTables)
library(psych)
library(jmv)
library(tidyverse)
```


#### **Load in Data**

```{r echo=TRUE}
#Load in the data frame into an object `dat`

dat <- read.csv("RegDiagnostics2.csv")

#View(dat)

```


#### **Change Missing Data Labels**

```{r echo=TRUE}
#Remove all `99` values from data frame and replace with `NA`s
# MAKE SURE YOU DO NOT HAVE AN ID SET TO 99 OR IT WILL REMOVE THIS CASE NUMBER!!

dat[dat=="99"] <- NA

```


### Exploratory Data Analysis

#### **Descriptive Statistics**
```{r echo=TRUE}
# Descriptive statistics  (Default: No Removal of Missing Data)
describe(dat)

# Descriptive statistics  (Removal of Missing Data)
describe(dat, 
         na.rm=FALSE)

```

#### **Generate Univariate and Bivariate Visualizations**

```{r warning = FALSE, message=FALSE, echo=TRUE}
# Univariate Histograms of Continuous Variables

ggplot(data = dat,
       mapping = aes(x = Amount)) +
  geom_histogram(bins = 10, color = "goldenrod3") +
  labs(y = "Count", x = "Amount")

ggplot(data = dat,
       mapping = aes(x = Need)) +
  geom_histogram(bins = 10, color = "goldenrod3") +
  labs(y = "Count", x = "Need")

ggplot(data = dat,
       mapping = aes(x = Interest)) +
  geom_histogram(bins = 10, color = "goldenrod3") +
  labs(y = "Count", x = "Interest")

ggplot(data = dat,
       mapping = aes(x = Happy)) +
  geom_histogram(bins = 10, color = "goldenrod3") +
  labs(y = "Count", x = "Happy")

ggplot(data = dat,
       mapping = aes(x = Belief)) +
  geom_histogram(bins = 10, color = "goldenrod3") +
  labs(y = "Count", x = "Belief")

```
```{r warning = FALSE, message=FALSE, echo=TRUE}
# Univariate Boxplots of Continuous Variables

ggplot(data = dat,
       mapping = aes(y = Amount)) +
  geom_boxplot() +
    labs(y = "Amount")

ggplot(data = dat,
       mapping = aes(y = Need)) +
  geom_boxplot() +
    labs(y = "Need")

ggplot(data = dat,
       mapping = aes(y = Interest)) +
  geom_boxplot() +
    labs(y = "Interest")

ggplot(data = dat,
       mapping = aes(y = Happy)) +
  geom_boxplot() +
    labs(y = "Happy")

ggplot(data = dat,
       mapping = aes(y = Belief)) +
  geom_boxplot() +
    labs(y = "Belief")

```
```{r warning=FALSE,message=FALSE,echo=TRUE}
# Bivariate Scatterplots of Amount with Predictors
ggplot(data = dat,
       mapping = aes(y = Amount, x = Need)) +
  geom_smooth(method = "lm", formula = y~x, se=FALSE) +
  geom_point() +
  labs(y = "Amount", x = "Need")

ggplot(data = dat,
       mapping = aes(y = Amount, x = Interest)) +
  geom_smooth(method = "lm", formula = y~x, se=FALSE) +
  geom_point() +
  labs(y = "Amount", x = "Interest")

ggplot(data = dat,
       mapping = aes(y = Amount, x = Belief)) +
  geom_smooth(method = "lm", formula = y~x, se=FALSE) +
  geom_point() +
  labs(y = "Amount", x = "Belief")

ggplot(data = dat,
       mapping = aes(y = Amount, x = Happy)) +
  geom_smooth(method = "lm", formula = y~x, se=FALSE) +
  geom_point() +
  labs(y = "Amount", x = "Happy")

```

#### **Generate Correlations Among Variables**
```{r echo=TRUE, warning=FALSE, message=FALSE}
# `psych` Correlation analyses (Listwise Deletion)
cor_listwise <- corr.test(dat[2:6], use="complete")
print(cor_listwise, short=FALSE)

# `jmv` Correlation analyses with Plots
corrMatrix(dat[2:6],
           ci = TRUE,
           plots = TRUE,
           plotDens = TRUE,
           plotStats = TRUE)
```

#### **Assess Missing Data/Pattern**
```{r echo=TRUE, warning=FALSE, message=FALSE}

# Calculate Amount of Missing Data
sum(is.na(dat[2:6]))/prod(dim(dat[2:6]))

# Assess Pattern of Missing Data
md.pattern(dat[2:6])

# Due to small amount of missing data we will use listwise deletion and proceed
```


#### **List-wise deletion of missing data**
```{r echo=TRUE, warning=FALSE, message=FALSE}
# Listwise deletion results in 7 cases removed from data
# Will create new data object to compare analyses later
dat_no_NA <- na.omit(dat)

```


#### **Assessment of Univariate Outliers**
```{r echo=TRUE, warning=FALSE,message=FALSE}
#Identify outliers
dat_no_NA[abs(scale(dat_no_NA$Belief)) > 3.29, ]
dat_no_NA[abs(scale(dat_no_NA$Need)) > 3.29, ]
dat_no_NA[abs(scale(dat_no_NA$Interest)) > 3.29, ]
dat_no_NA[abs(scale(dat_no_NA$Happy)) > 3.29, ]
dat_no_NA[abs(scale(dat_no_NA$Amount)) > 3.29, ]

#Belief has 2 univariate outliers (IDs 100 and 105)
#Need has 0
#Interest has 0
#Happy has 1 univariate outlier  (ID 100)
#Amount has 2 univariate outliers  (IDs 100 and 105)

```

#### **Remove Univariate Outliers**
```{r}
# Step needs to be conducted sequentially as shown below (can be expanded with larger datasets)
dat.no.uni1 <- dat_no_NA[!abs(scale(dat_no_NA$Belief)) > 3.29, ]
dat.no.uni2 <- dat.no.uni1[!abs(scale(dat.no.uni1$Need)) > 3.29, ]
dat.no.uni3 <- dat.no.uni2[!abs(scale(dat.no.uni2$Interest)) > 3.29, ]
dat.no.uni4 <- dat.no.uni3[!abs(scale(dat.no.uni3$Happy)) > 3.29, ]
dat_no_NA_UNI <- dat.no.uni4[!abs(scale(dat.no.uni4$Amount)) > 3.29, ]

```


#### **Reassess Univariate Normality After Univariate Outlier Removal**
```{r}
describe(dat_no_NA_UNI[2:6])
```


#### **Multivariate Outliers - Mahalanobis and Cook's Distances**
```{r}
# Mahalanobis Distance
dat_no_NA_UNI$mahal <- outlier(dat_no_NA_UNI[2:6])

dat_no_NA_UNI[abs(scale(dat_no_NA_UNI$mahal)) > 3.29, ]

## 2 Multivariate Outliers Identified (IDs 103 and 104)

#Cook's Distance
lm<-lm(Amount~Belief + Need + Interest + Happy, data = dat_no_NA_UNI)

#Generate Cook's Distance
dat_no_NA_UNI$cooks <- cooks.distance(lm)

dat_no_NA_UNI[abs(scale(dat_no_NA_UNI$cooks)) > 3.29, ]

## 1 Multivariate Outlier Identified (ID 103)

#Remove multivariate outliers
dat_no_NA_UMO <- dat_no_NA_UNI[!abs(scale(dat_no_NA_UNI$mahal)) > 3.29,]


```
#### **Multivariate Outlier Iteration Checking Code**
```{r}
# Mahalanobis Distance
dat_no_NA_UMO$mahal <- outlier(dat_no_NA_UMO[2:6])

dat_no_NA_UMO[abs(scale(dat_no_NA_UMO$mahal)) > 3.29, ]

# 0 Multivariate Outliers Identified

#Cook's Distance
lm<-lm(Amount~Belief + Need + Interest + Happy, data = dat_no_NA_UMO)

#Generate Cook's Distance
dat_no_NA_UMO$cooks <- cooks.distance(lm)

dat_no_NA_UMO[abs(scale(dat_no_NA_UMO$cooks)) > 3.29, ]

## Multivariate Outliers Identified IDs #1[142], #2[140,150], #3[111,136], #4[107,138], 
## #5[101,173], #6[113,131,162], #7[102,205], #8[197], #9[187], #10[166]

#Remove multivariate outliers
dat_no_NA_UMO <- dat_no_NA_UMO[!abs(scale(dat_no_NA_UMO$cooks)) > 3.29,]
```

#### **Rename Data Object for Additional Assumption Checks**
```{r}
dat_final <- dat_no_NA_UMO

# Remove extra datasets
rm(dat.no.uni1,
   dat.no.uni2,
   dat.no.uni3,
   dat.no.uni4,
   dat_no_NA,
   dat_no_NA_UMO,
   dat_no_NA_UNI)
```


#### **Multivariate Normality**
```{r}
# Mardia's Test of Multivariate Normality
mardia <- mvn(dat_final[2:6],
              mvnTest = "mardia",
              desc = FALSE)

# Henze-Zirkler's Test of Multivariate Normality
hz <- mvn(dat_final[2:6],
          mvnTest = "hz",
          desc = FALSE)

# Energy Test of Multivariate Normality
energy <- mvn(dat_final[2:6],
              mvnTest = "energy",
              desc = FALSE)

# Doornik-Hansen's Test of Multivariate Normality
dh <- mvn(dat_final[2:6],
          mvnTest = "dh",
          desc = FALSE)

mardia$multivariateNormality
hz$multivariateNormality
energy$multivariateNormality
dh$multivariateNormality

```

#### **Homoscedasticity**
```{r}

bplm<-lm(Amount~Belief+Need+Interest+Happy, data = dat_final[2:6])

#Breusch-Pagan test 
bptest(bplm, studentize=FALSE)
bptest(bplm, studentize=TRUE)

```


#### **Multicollinearity**
```{r}

#VIF and Tolerance
collin <- linReg(data = dat_final, 
                 dep = Amount, 
                 cov = c(Belief, Need, Interest, Happy),
                 blocks = list(list('Belief','Need','Interest','Happy')), 
                 r=FALSE,
                 r2=FALSE,
                 collin = TRUE)
collin

```


#### **Multiple Regression Cleaned vs. Uncleaned Data**
```{r}
corrMatrix(dat_final[2:6],
           ci = TRUE,
           plots = TRUE,
           plotDens = TRUE,
           plotStats = TRUE)

# Multiple regression not clean
linReg(data = dat,
       dep = 'Amount',
       covs = c('Belief', 'Need', 'Interest', 'Happy'),
       blocks = list(list('Belief', 'Need', 'Interest', 'Happy')),
       r2Adj = TRUE,
       modelTest=TRUE,
       ci = TRUE,
       stdEst = TRUE,
       ciStdEst = TRUE)

#Multiple regression clean
linReg(data = dat_final, 
       dep = 'Amount', 
       covs = c('Belief', 'Need', 'Interest', 'Happy'),
       blocks = list(list('Belief', 'Need', 'Interest', 'Happy')), 
       r2Adj = TRUE,
       modelTest=TRUE,
       ci = TRUE,
       stdEst = TRUE,
       ciStdEst = TRUE)


```

#### Model Pruning and Model Comparisons/Parsimonious Models
```{r}
#Simple Bivariate Regression (Amount ~ Belief)
lm1 <- lm(Amount~Belief, data = dat_final)

#Simple Standardized Bivariate Regression (Amount ~ Belief)
zlm1 <- lm(scale(Amount) ~ scale(Belief), data = dat_final)

#Call for standardized regression coefficients/summary
summary(zlm1)

#Call for standardized regression confidence intervals
confint(zlm1)
```
```{r}
#Multiple Linear Regression 
lm2 <- lm(Amount~Belief+Need, data = dat_final)

#Standardized Multiple Regression
zlm2 <- lm(scale(Amount) ~ scale(Belief) + scale(Need), data = dat_final)

#Call for standardized MR coefficients/summary
summary(zlm2)

#Call for standardized MR confidence intervals
confint(zlm2)
```
**Direct Model Comparison**
```{r}
#Model Comparison
anova(lm1, lm2)
```

#### **Run and Visualize Final Model**
```{r}
linReg(data = dat_final, 
       dep = 'Amount', 
       covs = c('Belief', 'Need'),
       blocks = list(list('Belief', 'Need')), 
       r2Adj = TRUE,
       modelTest=TRUE,
       ci = TRUE,
       stdEst = TRUE,
       ciStdEst = TRUE)
```
```{r}
# create predicted values from three predictors and save in object
dat_final$predictedF <- fitted(lm2)
dat_final$residuals <- resid(lm2)

# plot predicted line
ggplot(data = dat_final,
       mapping = aes(x = Amount, y = predictedF)) +
  geom_smooth(method = "lm", formula = y ~ x) +
  geom_point() +
  labs(x = "Amount Given", y = "Predicted Amount Given")

```

**Checking residual plots**
```{r}
modelF <- linReg(data = dat_final, 
                 dep = Amount, 
                 covs = vars(Belief,Need),
                 blocks = list(list('Belief','Need')), 
                 modelTest = FALSE,
                 r=FALSE,
                 r2=FALSE,
                 resPlots=TRUE)
modelF

ggplot(data = dat_final,
       mapping = aes(x = residuals)) +
  geom_histogram()
```

