---
title: "Confirmatory Structure Analysis Demonstration"
author: "Professor Christopher S. Schmank"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning = FALSE}
library(lavaan)
library(lavaanPlot)
library(jmv)
library(psych)
```

#### Load in Data

```{r}
wiscsem <- read.csv("wiscsem.csv")[,-c(1:2,13)]
cogability <- read.csv("cogability.csv")[,c(2:10)]
```

#### Descriptive Statistics

```{r}
describe(wiscsem)
describe(cogability)
```

#### `CorrMatrix()` output

```{r}
corrMatrix(wiscsem,plots = TRUE,plotDens = TRUE,plotStats = TRUE)$plot
corrMatrix(cogability,plots = TRUE,plotDens = TRUE,plotStats = TRUE)$plot
```

### Confirmatory Factor Analysis (CFA)

#### Model Specification (i.e., Measurement Model)

One-factor Model:
```{r}
onefactor <- ' # Latent variable
g =~ info + comp + arith + simil + vocab + digit + pictcomp + parang + block + object

# Residuals for manifest variables
info~~info
comp~~comp
arith~~arith
simil~~simil
vocab~~vocab
digit~~digit
pictcomp~~pictcomp
parang~~parang
block~~block
object~~object

# Variances for latent variables
g~~1*g
'

```

Two-factor (Uncorrelated) Model:
```{r}
twofactor <- ' # Latent variables
verbal =~ info + comp + arith + simil + vocab + digit
performance =~ pictcomp + parang + block + object

# Residuals for manifest variables
info~~info
comp~~comp
arith~~arith
simil~~simil
vocab~~vocab
digit~~digit
pictcomp~~pictcomp
parang~~parang
block~~block
object~~object

# Variances for latent variables
verbal~~1*verbal
performance~~1*performance
'

```

Two-factor (Correlated) Model:
```{r}
ctwofactor <- ' # Latent variables
verbal =~ info + comp + arith + simil + vocab + digit
performance =~ pictcomp + parang + block + object

# Latent variable correlation
verbal ~~ performance

# Residuals for manifest variables
info~~info
comp~~comp
arith~~arith
simil~~simil
vocab~~vocab
digit~~digit
pictcomp~~pictcomp
parang~~parang
block~~block
object~~object

# Variances for latent variables
verbal~~1*verbal
performance~~1*performance
'

```

## Run the model, summarize the parameter estimates, and plot: 

```{r message=FALSE, warning=FALSE}
# Run the model: One Factor Model
fit <- lavaan(onefactor, 
              data = wiscsem, 
              std.lv = TRUE)

# Get parameter estiamtes and fit statistics 
summary(fit, 
        fit.measures = TRUE, 
        standardized = TRUE)

lavaanPlot(fit, 
           coefs = TRUE, 
           stand = TRUE)
```

```{r message=FALSE, warning=FALSE}
# Run the model: Two Factor Model
fit2 <- lavaan(twofactor, 
               data = wiscsem, 
               std.lv = TRUE)

# Get parameter estiamtes and fit statistics 
summary(fit2, 
        fit.measures = TRUE, 
        standardized = TRUE)

lavaanPlot(fit2, 
           coefs = TRUE, 
           stand = TRUE)
```

```{r message=FALSE, warning=FALSE}
# Run the model: Two Factor Model
fit2c <- lavaan(ctwofactor, 
                data = wiscsem, 
                std.lv = TRUE)

# Get parameter estiamtes and fit statistics 
summary(fit2c, 
        fit.measures = TRUE, 
        standardized = TRUE)

lavaanPlot(fit2c, 
           coefs = TRUE, 
           stand = TRUE,
           covs = TRUE)
```

## Direct Model Fit Comparisons (Chi-square Difference Test)

```{r}
# Can compare fit of our 2-factor models because they are nested (same items) AND one has fewer degrees of freedom
anova(fit2,fit2c)

anova(fit, fit2c)

# Then we can call for the summary of the "best" model and the model fit indices
summary(fit2c, 
        fit.measures = TRUE, 
        standardized = TRUE)

fitMeasures(fit2c)
```

### Structural Equation Model

#### Model Specification (Step 1 SEM: Measurement Model)

Correlated Three Factor Model:
```{r}
cthreefactor <- ' # Latent variables
Fluid =~ V1 + V2 + V3
Verbal =~ V4 + V5 + V6
Visuospatial =~ V7 + V8 + V9

# Latent variable correlation
Fluid ~~ Verbal
Fluid ~~ Visuospatial
Verbal ~~ Visuospatial

# Residuals for manifest variables
V1 ~~ V1
V2 ~~ V2
V3 ~~ V3
V4 ~~ V4
V5 ~~ V5
V6 ~~ V6
V7 ~~ V7
V8 ~~ V8
V9 ~~ V9

# Variances for latent variables
Fluid~~1*Fluid
Verbal~~1*Verbal
Visuospatial~~1*Visuospatial
'
```

## Model Specification (Step 2 SEM: Full Structural Model)

Fluid Intelligence Predicts Verbal/Visuospatial Model:
```{r}
gfpredicts <- ' # Latent variables
Fluid =~ V1 + V2 + V3
Verbal =~ V4 + V5 + V6
Visuospatial =~ V7 + V8 + V9

# Regressions

Verbal + Visuospatial ~ Fluid

# Residuals for manifest variables
V1 ~~ V1
V2 ~~ V2
V3 ~~ V3
V4 ~~ V4
V5 ~~ V5
V6 ~~ V6
V7 ~~ V7
V8 ~~ V8
V9 ~~ V9

# Variances for latent variables
Fluid~~1*Fluid
Verbal~~1*Verbal
Visuospatial~~1*Visuospatial
'
```

Fluid Intelligence Predicts Verbal/Visuospatial Correlated Model:
```{r}
gfpredictsc <- ' # Latent variables
Fluid =~ V1 + V2 + V3
Verbal =~ V4 + V5 + V6
Visuospatial =~ V7 + V8 + V9

# Correlated Latent variables
Verbal ~~ Visuospatial

# Regressions

Verbal + Visuospatial ~ Fluid

# Residuals for manifest variables
V1 ~~ V1
V2 ~~ V2
V3 ~~ V3
V4 ~~ V4
V5 ~~ V5
V6 ~~ V6
V7 ~~ V7
V8 ~~ V8
V9 ~~ V9

# Variances for latent variables
Fluid~~1*Fluid
Verbal~~1*Verbal
Visuospatial~~1*Visuospatial
'
```

## Run the CFA measurement model, summarize the parameter estimates, and plot 

```{r message=FALSE, warning=FALSE}
# Run the model: Three Factor Model
cfafit <- lavaan(cthreefactor, 
                 data = cogability, 
                 std.lv = TRUE)

# Get parameter estiamtes and fit statistics 
summary(cfafit, 
        fit.measures = TRUE, 
        standardized = TRUE)

lavaanPlot(cfafit, 
           coefs = TRUE, 
           stand = TRUE,
           covs = TRUE)
```

## Run the SEM model(s), summarize the parameter estimates, and plot 

```{r message=FALSE, warning=FALSE}
# Run the model: Gf Predicts SEM
semfit <- lavaan(gfpredicts, 
                 data = cogability, 
                 std.lv = TRUE)

# Get parameter estiamtes and fit statistics 
summary(semfit, 
        fit.measures = TRUE, 
        standardized = TRUE)

lavaanPlot(semfit, 
           coefs = TRUE, 
           stand = TRUE,
           covs = TRUE)
```


```{r message=FALSE, warning=FALSE}
# Run the model: Gf Predicts Correlated SEM
semfitc <- lavaan(gfpredictsc, 
                 data = cogability, 
                 std.lv = TRUE)

# Get parameter estiamtes and fit statistics 
summary(semfitc, 
        fit.measures = TRUE, 
        standardized = TRUE)

lavaanPlot(semfitc, 
           coefs = TRUE, 
           stand = TRUE,
           covs = TRUE)
```

#### Direct Model Fit Comparisons (Chi-square Difference Test)

```{r}
# Can compare fit of our 2-factor models because they are nested (same items) AND one has fewer degrees of freedom
anova(semfit,semfitc)
```

